Icicle: write once, run once

### 2016-05-25 2016fhpc/section/01-Introduction.tex
Wibbles

At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple queries over hundreds of gigabytes worth of data.\footnote{In 2016 this was a lot of data.} For such applications we must automatically fuse these separate queries and be sure that the result can be executed in a single pass over the input. We also ingest multiple gigabytes of new data per day, and must incrementally update existing features without recomputing them all from scratch. Our feature generation process is performed in parallel on hundreds of nodes of a cloud based system, and if we performed neither fusion or incremental update then the cost of the computation would begin to exceed the salaries of the developers.

@@

A joint query like this can be converted to a back-end language such a SQL, but doing it by hand it tedious and error prone. As the three sub queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter. We instead need to lift each predicate to an expression-level conditional:
%
\begin{code}
  SELECT COUNT(IF(open > close, 1, 0))
       , COUNT(IF(open < close, 1, 0))
       , SUM  (IF(open > close, open, 0))
       / COUNT(IF(open > close, 1, 0))
  FROM stocks;
\end{code}

As we see, the result of query fusion also tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.

Joint queries such as the stocks example can be evaluated in a streaming, incremental fashion. As a counter example, suppose we have a table with two fields @key@ and @value@, and we wish to find the mean of values whose key matches the last one in the table. We might try something like:
%
\begin{code}
  table kvs { key : Int; value : Int }
  query avg = let k = last key
              in  filter (key == k) in mean value;
\end{code}

Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} for doing so implied by the above query cannot be evaluated in a streaming fashion. When we are streaming through the table we always have access to the last key in the stream, but to perform the filter to select rows that match this key would require that we stream through the table again from the start. We need a better solution. With this in mind, the contributions of this paper are:
\begin{itemize}
\item
we present a type system that guarantees any two queries on the same input table can be fused, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};

@@

>>>>

At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple queries over hundreds of gigabytes worth of data.\footnote{In 2016 this was a lot of data.} For such applications we must automatically fuse these separate queries and be sure that the result can be executed in a single pass over the input. We also ingest multiple gigabytes of new data per day, and must incrementally update existing features without recomputing them all from scratch. Our feature generation process is executed in parallel on hundreds of nodes on a cloud based system, and if we performed neither fusion or incremental update then the cost of the computation would begin to exceed the salaries of the developers.

@@

A joint query like this can be converted to a back-end language such a SQL, but doing it by hand it tedious and error prone. As the three sub queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter. We instead need to lift each predicate to an expression-level conditional:
\begin{code}
  SELECT COUNT(IF(open > close, 1, 0))
       , COUNT(IF(open < close, 1, 0))
       , SUM  (IF(open > close, open, 0))
       / COUNT(IF(open > close, 1, 0))
  FROM stocks;
\end{code}
As we see, the result of query fusion also tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.

Joint queries such as the stocks example can be evaluated in a streaming, incremental fashion. As a counter example, suppose we have a table with two fields @key@ and @value@, and we wish to find the mean of values whose key matches the last one in the table. We might try something like:
\begin{code}
  table kvs { key : Int; value : Int }
  query avg = let k = last key
              in  filter (key == k) in mean value;
\end{code}
Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in a streaming fashion. While streaming through the table we always have access to the last key in the stream, but to perform a filter to select rows that match this key would require that we stream through the table again from the start. We need a better solution. With this in mind, the contributions of this paper are:
\begin{itemize}
\item
we present a type system that guarantees any two queries on the same input table can be fused, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};

@@

\TODO{Touch on related work}
